Learning action models from {\em state-invariants}

### 2019-02-14 doc/ijcai-19-deductivelearning/deductivelearning.tex
first full version (no experiments though) ijcai19 paper on deductive learning

%%%%%%%%%%%%%%%%%% Added for this paper
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{listings}% http://ctan.org/pkg/listings
\lstset{
  basicstyle=\ttfamily,
  mathescape
}
\usepackage{ wasysym }
\newcommand{\tup}[1]{{\langle #1 \rangle}}
\newcommand{\pre}{\mathsf{pre}}     % precondition
\newcommand{\del}{\mathsf{del}}     % effect
\newcommand{\add}{\mathsf{add}}     % effect
\newcommand{\eff}{\mathsf{eff}}     % effect
\newcommand{\cond}{\mathsf{cond}}   % conditional effect
\newcommand{\true}{\mathsf{true}}   % true
\newcommand{\false}{\mathsf{false}} % false
\newcommand{\PE}{\mathrm{PE}}     % precondition
\newcommand{\strips}{\textsc{Strips}}

@@

@@

{\em State invariants} is a kind of state-constraints useful for computing more compact state representations~\cite{helmert2009concise} or making {\em satisfiability planning} and {\em backward search} more efficient~\cite{rintanen2014madagascar,alcazar2015reminder}. Given a classical planning problem $P=\tup{F,A,I,G}$, a {\em state invariant} is a formula $\phi$ that holds at the initial state of a given classical planning problem, $I\models \phi$, and at every state $s$, built from $F$, that is reachable from $I$ by applying actions in $A$. For instance, Figure~\ref{fig:strongest-invariant} shows five clauses that define {\em state invariants} for the {\em blocksworld} planning domain.

@@

A {\em mutex} (mutually exclusive) is a state invariant that takes the form of a binary clause and indicates a pair of different properties that cannot be simultaneously true~\cite{kautz:mutex:IJCAI1999}. For instance in a three-block {\em blocksworld}, $\neg on(block_A,block_B)\vee \neg on(block_A,block_C)$ is a {\em mutex} because $block_A$ can only be on top of a single block.

A {\em domain invariant} is an instance-independent state-invariant, i.e. holds for any possible initial state and any possible set of objects. Therefore, if a given state $s$ holds $s\nvDash \phi$ such that $\phi$ is a {\em domain invariant}, it means that $s$ is not a valid state. Domain invariants are often compactly defined as {\em lifted invariants} (also called schematic invariants)~\cite{rintanen:schematicInvariants:AAAI2017}.

@@

\subsection{The sampling space}
We define a {\em learning example} as a sequence $\mathcal{O}=\tup{s_0^o,s_1^o \ldots, s_m^o}$ of partial states, except for the initial state $s_0^o$, which is fully observed. Intermediate states can be missing meaning that are {\em unobserved}, so transiting between two consecutive observed states in $\mathcal{O}$ may require the execution of more than a single action ($\theta(s_i^o,\tup{a_1,\ldots,a_k})=s_{i+1}^o$, where ${\small k\geq 1}$ is unknown but finite. A partially observed state $s_i^o$ is one in which $|s_i^o| < |F|$; i.e., a state in which at least a fluent of $F$ is not observable. Note that this definition also comprises the case $|s_i^o| = 0$, when the state is fully unobservable.

@@

Despite any element of ${\mathcal I}_{\Psi,\xi}$ can {\em a priori} appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of schema $\xi$, in practice the actual space of possible \strips\ schemata is bounded by constraints of two kinds:
\begin{enumerate}
\item {\bf Syntactic constraints}. \strips\ constraints require $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$. Considering exclusively these syntactic constraints, the size of the space of possible \strips\ schemata is given by $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$. {\em Typing constraints} are also of this kind~\cite{mcdermott1998pddl}. 
\item {\bf Observation constraints}. The observation of the actions and states resulting from the execution of a plan depicts {\em semantic knowledge} that constraints further the space of possible action schemata.   
\end{enumerate}

@@

One can also introduce {\em domain-specific knowledge} to constrain further the space of possible schemata. For instance, in the {\em blocksworld} one can argue that {\small\tt on($v_1$,$v_1$)} and {\small\tt on($v_2$,$v_2$)} will not appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists of an action schema $\xi$ because, in this specific domain, a block cannot be on top of itself. {\it State invariants} are {\em domain-specific knowledge} that can be seen either as {\em syntactic} or {\em semantic} constraints. On the one hand, {\it state invariants} constrain the space of possible action models but on the other hand, they can complete partial observations of the states traversed by a plan.

\subsection{The learning task}
We define the task of learning a planning action model from {\em state-invariants} as a tuple $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$, where:
\begin{itemize}
\item $\mathcal{M}$ is the {\em initial empty model} that contains only the name, $name(\xi)$, and parameters, $pars(\xi)$, of each action model $\xi$ to be learned.
\item $\mathcal{O}=\tup{s_0^o,s_1^o \ldots, s_m^o}$ is a single learning example. This example can be reduced to its minimal expression $\mathcal{O}^*=\tup{s_0^o, s_m^o}$ comprissing at least two state observations, a full initial state $s_0^o$ and a partially observed state $s_m^o$. The set of predicates $\Psi$ that shape the setf of fluents $F$ is deducible from $\mathcal{O}$ since $s_0^o$ is a fully observed state.
\item $\Phi$ is a set of {\em state-invariants} that define constraints about the set of possible states.
\end{itemize}

Next section shows how to exploit our compact encoding of \strips\ action models to solve a $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task with an off-the-shelf classical planner.

\section{Learning action models from {\em state-invariants} with classical planning}
\label{sec:compilation}

Our sampling space follows the {\em open world} assumption, i.e. what is not observed is considered unknown. {\em State-invariants} are helpful to infer new knowledge that was unobserved in the learning examples. Given a {\em domain mutex} $\tup{p,q}$ and a state observation $s_j^o\in \mathcal{O}(\tau)$, {\small $(1\leq j\leq m)$}, such that literal $p(\omega)\in s_j^o$ is an instantiation of predicate $p$ over some subset of objects $\omega\subseteq\Omega^{|pars(p)|}$ then, the state observation $s_j^o$ can be safely completed adding the new literal $\neg q(\omega)$ (despite $\neg q(\omega)$ was actually unobserved). For instance, if the literal {\tt\small holding(blockA)} is observed in a particular blocksword state and we know the {\em domain mutex} $\forall x\ holding(x)\leftrightarrow\neg clear(x)$ we can safely extend the observation with literal {\tt\small $\neg$clear(blockA)} (despite this literal was actually unobserved).

Given a $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task we build and solve a classical planning problem $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I,G_{\Lambda}}$ such that:
\begin{itemize}
\item $F_{\Lambda}$ extends $F$ with a fluent {\small$mode_{inval}$}, to indicate whether an action model is {\em inconsistent} with the input {\em state-invariants} $\Phi$, a fluent {\small$mode_{insert}$}, to indicate whether action models are being programmed, and the fluents for the propositional encoding of the corresponding space of STRIPS action models. As explained, this is a set of fluents of the type $\{pre\_e\_\xi, eff\_e\_\xi\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}$.

@@

\item $A_{\Lambda}$ replaces the actions in $A$ with two types of actions.
\begin{enumerate}
\item Actions for {\em inserting} a {\em precondition}, {\em positive} effect or {\em negative} effect in $\xi$ following the syntactic constraints of \strips\ models. 
\begin{itemize}
\item Actions which support the addition of a {\em precondition} $p\in {\mathcal I}_{\Psi,\xi}$ to the action model $\xi$. A precondition $p$ is inserted in $\xi$ when neither $pre_p$, $eff_p$ exist in $\xi$.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertPre_{p,\xi}})=&\{\neg pre\_p\_\xi, \neg eff\_p\_\xi, mode_{insert}\},\\
\cond(\mathsf{insertPre_{p,\xi}})=&\{\emptyset\}\rhd\{pre\_p\_\xi\}.
\end{align*}
\end{small}

\item Actions which support the addition of a {\em negative} or {\em positive} effect $p\in {\mathcal I}_{\Psi,\xi}$ to the action model $\xi$.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertEff_{p,\xi}})=&\{\neg eff\_p\_\xi, mode_{insert}\},\\
\cond(\mathsf{insertEff_{p,\xi}})=&\{\emptyset\}\rhd\{eff\_p\_\xi\}.
\end{align*}
\end{small}
\end{itemize}

\item Actions for {\em applying} an action model $\xi$ built by the {\em insert} actions and bounded to objects $\omega\subseteq\Omega^{|pars(\xi)|}$ (where $\Omega$ is the set of {\em objects} used to induce the fluents $F$ by assigning objects in $\Omega$ to the $\Psi$ predicates, and $\Omega^k$ is the $k$-th Cartesian power of $\Omega$). The action parameters, $pars(\xi)$, are bound to the objects in $\omega$ that appear in the same position. These actions validate also that any state traversed by $P_{\Lambda}$ solutions is {\em consistent} with the {\em state-invariants} $\Phi$. The definition $\mathsf{apply_{\xi,\omega}}$ actions is also more compact in our compilation that the one previously proposed by~\citeauthor{aineto2018learning}~\citeyear{aineto2018learning} since are not using disjunctions to code the possible preconditions of an action schema.
\end{enumerate}
\end{itemize}

\begin{small}
\begin{align*}
\pre(\mathsf{apply_{\xi,\omega}})=&\{\neg mode_{inval}\},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{pre\_p\_\xi\wedge eff\_p\_\xi\}\rhd\{\neg p(\omega)\}_{\forall p\in{\mathcal I}_{\Psi,\xi}},\\
&\{\neg pre\_p\_\xi \wedge eff\_p\_\xi\}\rhd\{p(\omega)\}_{\forall p\in{\mathcal I}_{\Psi,\xi}}\},\\
&\{pre\_p\_\xi \wedge \neg p(\omega)\}\rhd\{mode_{inval}\}_{\forall p\in{\mathcal I}_{\Psi,\xi}},\\
&\{\neg\phi\}\rhd\{mode_{inval}\}_{\forall \phi\in\Phi},\\
&\{\emptyset\}\rhd\{\neg mode_{insert}\},
\end{align*}
\end{small}

\subsection{Pruning inconsistent action models with {\em domain mutex}}
We define a {\em domain mutex} as a $\tup{p,q}$ predicates pair where both $p\in\Psi$ and $q\in\Psi$ are predicates that shape the set of fluents $F$ of a given planning problem and such that they satisfy the following formulae $p\rightarrow \neg q$ where the predicate variables are universally quantified. For instance, predicates $holding(x)$ and $clear(x)$ from the {\em blocksworld} are {\em domain mutex} while predicates $clear(x)$ and $ontable(x)$ are not ($\forall x\ clear(x)\leftrightarrow\neg ontable(x)$ does not hold for every possible {\em blocksworld} state). We pay special attention to {\em domain mutex} because they identify the {\em properties} of a given type of objects~\cite{fox:TIM:JAIR1998} and because they enable (1) effectively pruning of inconsistent \strips\ action models and (2) effective completion of partially observed states.

Our approach to implement this pruning is extending the conditional effects of the $\mathsf{insertPre_{p,\xi}}$ and $\mathsf{insertPre_{p,\xi}}$ actions (i.e., the actions that determine a solution model $\mathcal{M}$) with extra conditional effects indicating that the programmed model is {\em invalid} (i.e., inconsistent with a {\em domain mutex} in $\Phi$). Note that this {\em consistency} checking is more effective than the one implemented at the $\mathsf{apply_{\xi,\omega}}$ actions since $\mathsf{insertPre_{p,\xi}}$ and $\mathsf{insertPre_{p,\xi}}$ actions appear at an earlier stage of the planning process.

Formally, given a {\em domain mutex} $(p,q)$, s.t. both $p$ and $q$ belong to $\in{\mathcal I}_{\Psi,\xi}$, we extend the actions for setting a precondition $p$ in a given action schema $\xi$ as follows:

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertPre_{p,\xi}})=&\{\neg pre_{p}(\xi), \neg eff_{p}(\xi),\\
&mode_{insert},\neg mode_{inval}\},\\
\cond(\mathsf{insertPre_{p,\xi}})=&\{\emptyset\}\rhd\{pre_{p}(\xi)\},\\
&\{pre_{q}(\xi)\}\rhd\{mode_{inval}\}.
\end{align*}
\end{small}
The same procedure is applied for action $insertPre_{q,\xi}$ to ban programming precondition $q$ iff $pre_{p}(\xi)$ precondition is already set. A similar procedure is also applied to $\mathsf{insertEff_{p,\xi}}$ and $\mathsf{insertEff_{q,\xi}}$ actions for banning in this case, two {\em negative effects} (or two {\em positive effects}) that are {\em domain mutex}. Now we show the actions that ban programming a positive (or negative) $p$ effect if its corresponding $q$ effect is already programmed:

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertEff_{p,\xi}})=&\{\neg eff_{p}(\xi), mode_{insert},\neg mode_{inval}\},\\
\cond(\mathsf{insertEff_{p,\xi}})=&\{\emptyset\}\rhd\{eff_{p}(\xi),\\
&\{pre_{q}(\xi),eff_{q}(\xi),pre_{p}(\xi)\}\rhd\{mode_{inval}\},\\
&\{\neg pre_{q}(\xi),eff_{q}(\xi),\neg pre_{p}(\xi)\}\rhd\{mode_{inval}\}.
\end{align*}
\end{small}

\subsection{The bias of the initially {\em empty} action model}
Classical planners tend to preffer shorter solution plans, so our compilation may introduce a bias to $P=\tup{F,A[\cdot],I,G}$ problems preferring solutions that are referred to action models with a shorter number of {\em preconditions}/{\em effects}. In more detail, all $\{pre\_e\_\xi, eff\_e\_\xi\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}$ fluents are false at the initial state of our $P'=\tup{F',A',I,G}$ compilation so classical planners tend to solve $P'$ with plans that require a shorter number of {\em insert} actions.

This bias could be eliminated defining a cost function for the actions in $A'$ (e.g. {\em insert} actions has {\em zero cost} while $\mathsf{apply_{\xi,\omega}}$ actions has a {\em positive constant cost}). In practice we use a different approach to disregard the cost of {\em insert} actions because classical planners are not proficiency optimizing {\em plan cost} with zero-cost actions. Instead, our approach is to use a SAT-based planner~\cite{rintanen2014madagascar} because it can apply all actions for inserting preconditions in a single planning step (these actions do not interact). Further, the actions for inserting action effects are also applied in a single planning step so the plan horizon for programming any action model is always bound to 2, which significantly reduces the planning horizon.

Our compilation for {\em planning with unknown domain models} can then be understood as an extension of the SATPLAN approach for classical planning~\cite{kautz1992planning} with two additional initial layers: a first layer for inserting the action preconditions and a second one for inserting the action effects. These two extra layers are followed by the typical $N$ layers of the SATPLAN encoding (extended however to apply the action models that are determined by the previous two initial layers, the $\mathsf{apply_{\xi,\omega}}$ actions). Regarding again the example of Figure~\ref{fig:plan-lplan}, this means that steps [00-04] are applied in paralel in the first SATPLAN layer, steps [05-13] are applied in paralel in the second layer and each step [14-17] is applied sequentially and correponds to a differerent SATPLAN layer (so just six layers are necesary to compute the example plan of Figure~\ref{fig:plan-lplan}).

The SAT-based planning approach is also convenient because its ability to deal with classical planning problems populated with dead-ends and because symmetries in the insertion of preconditions/effects into an action model do not affect to the planning performance.

\subsection{Compilation properties}
\begin{mylemma}
Soundness. Any classical plan $\pi_{\Lambda}$ that solves $P_{\Lambda}$ produces a \strips\ model $\mathcal{M}$ that solves the $\Lambda=\tup{P,\Phi}$ learning task.
\end{mylemma}

>>>>

%%%%%%%%%%%%%%%%%% Added for this paper
\usepackage{enumitem}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{listings}% http://ctan.org/pkg/listings
\lstset{
  basicstyle=\ttfamily,
  mathescape
}
\usepackage{ wasysym }
\newcommand{\tup}[1]{{\langle #1 \rangle}}
\newcommand{\pre}{\mathsf{pre}}     % precondition
\newcommand{\del}{\mathsf{del}}     % effect
\newcommand{\add}{\mathsf{add}}     % effect
\newcommand{\eff}{\mathsf{eff}}     % effect
\newcommand{\cond}{\mathsf{cond}}   % conditional effect
\newcommand{\true}{\mathsf{true}}   % true
\newcommand{\false}{\mathsf{false}} % false
\newcommand{\PE}{\mathrm{PE}}     % precondition
\newcommand{\strips}{\textsc{Strips}}

@@

%The {\em classical planning compilation} for the learning \strips\ action models~\cite{aineto2018learning} can be understood as an extension of the SATPLAN approach for classical planning~\cite{kautz1992planning} with two additional initial layers: a first layer for inserting the action preconditions and a second one for inserting the action effects. These two extra layers are followed by the typical $N$ layers of the SATPLAN encoding (extended however to apply the action models that are determined by the previous two initial layers). Regarding again the example of Figure~\ref{fig:plan-lplan}, this means that steps [00-04] are applied in paralel in the first SATPLAN layer, steps [05-13] are applied in paralel in the second layer and each step [14-17] is applied sequentially and correponds to a differerent SATPLAN layer (so just six layers are necesary to compute the example plan of Figure~\ref{fig:plan-lplan}).

@@

{\em State-invariants} is a kind of state-constraints useful for computing more compact state representations~\cite{helmert2009concise} or making {\em satisfiability planning} and {\em backward search} more efficient~\cite{rintanen2014madagascar,alcazar2015reminder}. Given a classical planning problem $P=\tup{F,A,I,G}$, a {\em state-invariant} is a formula $\phi$ that holds at the initial state of a given classical planning problem, $I\models \phi$, and at every state $s$, built from $F$, that is reachable from $I$ by applying actions in $A$. For instance, Figure~\ref{fig:strongest-invariant} shows five clauses that define {\em state-invariants} for the {\em blocksworld} planning domain.

@@

A {\em mutex} (mutually exclusive) is a {\em state-invariant} that takes the form of a binary clause and indicates a pair of different properties that cannot be simultaneously true~\cite{kautz:mutex:IJCAI1999}. For instance in a three-block {\em blocksworld}, $\neg on(block_A,block_B)\vee \neg on(block_A,block_C)$ is a {\em mutex} because $block_A$ can only be on top of a single block.

A {\em domain invariant} is an instance-independent state-invariant, i.e. holds for any possible initial state and any possible set of objects. Therefore, if a given state $s$ holds $s\nvDash \phi$ such that $\phi$ is a {\em domain invariant}, it means that $s$ is not a valid state. Domain invariants are often compactly defined as {\em lifted invariants} (also called schematic invariants~\cite{rintanen:schematicInvariants:AAAI2017}).

@@

\subsection{The sampling space}
We define a {\em learning example} as a sequence $\mathcal{O}=\tup{s_0^o,s_1^o \ldots, s_m^o}$ of partial states, except for the initial state $s_0^o$ which is fully observed. Intermediate states can be missing meaning that are {\em unobserved}, so transiting between two consecutive observed states in $\mathcal{O}$ may require the execution of more than a single action ($\theta(s_i^o,\tup{a_1,\ldots,a_k})=s_{i+1}^o$, where ${\small k\geq 1}$ is unknown but finite. A partially observed state $s_i^o$, ${\small 1\leq i\leq m}$, is one in which $|s_i^o| < |F|$; i.e., a state in which at least a fluent of $F$ is not observable. Note that this definition also comprises the case $|s_i^o| = 0$, when the state is fully unobservable.

@@

Despite any element of ${\mathcal I}_{\Psi,\xi}$ can {\em a priori} appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of schema $\xi$, in practice the actual space of possible \strips\ schemata is bounded by:
\begin{enumerate}
\item {\bf Syntactic constraints}. \strips\ constraints require $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$. Considering exclusively these syntactic constraints, the size of the space of possible \strips\ schemata is given by $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$. {\em Typing constraints} are also of this kind~\cite{mcdermott1998pddl}. 
\item {\bf Observation constraints}. The {\em learning examples}, that are observations of plan executions, depict {\em semantic knowledge} that constraints further the space of possible action schemata.   
\end{enumerate}

@@

One can also introduce {\em domain-specific knowledge} to constrain further the space of possible schemata. For instance, in the {\em blocksworld} one can argue that {\small\tt on($v_1$,$v_1$)} and {\small\tt on($v_2$,$v_2$)} will not appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists of an action schema $\xi$ because, in this specific domain, a block cannot be on top of itself. {\it State-invariants} are {\em domain-specific knowledge} that can be used to define new {\em syntactic} and {\em semantic} constraints over the space of possible action schemata.

\subsection{The learning task}
We define the task of learning a planning action model from {\em state-invariants} as a tuple $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$, where:
\begin{itemize}
\item $\mathcal{M}$ is the {\em initial empty model} that contains only the name, $name(\xi)$, and parameters, $pars(\xi)$, of each action model $\xi$ to be learned.
\item $\mathcal{O}=\tup{s_0^o,s_1^o \ldots, s_m^o}$ is a single learning example. This example can be reduced to its minimal expression $\mathcal{O}^*=\tup{s_0^o, s_m^o}$ comprissing at least two state observations, a full initial state $s_0^o$ and a partially observed state $s_m^o$. The set of predicates $\Psi$ and objects $\Omega$ that shape the set of fluents $F$ is deducible from $\mathcal{O}$ since $s_0^o$ is a fully observed state.
\item $\Phi$ is a set of {\em state-invariants} that define constraints about the set of possible states.
\end{itemize}

A {\em solution} to a learning task $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ is a model $\mathcal{M}'$ that is consistent with the the {\em initial empty model} $\mathcal{M}$, the learning example $\mathcal{O}$ and the set of {\em state invariants} in $\Phi$.

\section{Learning action models from {\em state-invariants} with classical planning}
\label{sec:compilation}
This section shows how to exploit our compact encoding of \strips\ action models to solve a $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task with an off-the-shelf classical planner.

\subsection{Completing partially observed states with {\em schematic mutex}}
Our sampling space follows the {\em open world} assumption, i.e. what is not observed is considered unknown. Here we describe a pre-processing mechanism to add new knowledge that completes states that are partially observed in a $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task using a particular case of {\em state-invariants} that we call {\em schematic mutex}. We pay special attention to {\em schematic mutex} because they identify the {\em properties} of a given type of objects~\cite{fox:TIM:JAIR1998} and because they enable (1) effectively pruning of inconsistent \strips\ action models and (2) effective completion of partially observed states.

We define a {\em schematic mutex} as a $\tup{p,q}$ predicates pair where both $p\in\Psi$ and $q\in\Psi$ are predicates that shape the set of state variables $F$ and such that they satisfy the following formulae $p\rightarrow \neg q$, where the predicate variables are universally quantified. For instance, predicates $holding(v_1)$ and $clear(v_1)$ from the {\em blocksworld} are {\em schematic mutex} while predicates $clear(v_1)$ and $ontable(v_1)$ are not because $\forall v_1\ clear(v_1)\leftrightarrow\neg ontable(v_1)$ does not hold for every possible {\em blocksworld} state.

Given a {\em schematic mutex} $\tup{p,q}$ and a state observation $s_j^o\in \mathcal{O}$, {\small $(1\leq j\leq m)$}, such that $p(\omega)\in s_j^o$ is a literal built instantiating predicate $p$ with some subset of objects $\omega\subseteq\Omega^{|pars(p)|}$ then, the state observation $s_j^o$ can be safely completed adding the new literal $\neg q(\omega)$ (despite $\neg q(\omega)$ was actually unobserved). For instance, if the literal {\tt\small holding(blockA)} is observed in a particular blocksword state and we know the {\em schematic mutex} $\forall x\ holding(x)\rightarrow\neg clear(x)$ we can safely extend that state observation with literal {\tt\small $\neg$clear(blockA)} (despite this literal was actually unobserved).

\subsection{Pruning inconsistent action models with {\em schematic mutex}}
Our approach to implement this pruning is adding new conditional effects to the classical planning compilation for the learning of \strips\ action models~\cite{aineto2018learning} that capture when the programmed model is inconsistent with a {\em schematic mutex} in $\Phi$. Given a $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task we build and solve a classical planning problem $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I,G_{\Lambda}}$ that extends the original compilation as follows:
\begin{itemize}
\item $F_{\Lambda}$ extends $F$ with a fluent {\small$mode_{inval}$}, to indicate whether an action model is {\em inconsistent} with the input {\em state-invariants} $\Phi$. Like the original compilation we define a fluent {\small$mode_{insert}$}, to indicate whether action models are being programmed, and the fluents for the propositional encoding of the corresponding space of STRIPS action models. As explained, this is a set of fluents of the type $\{pre\_e\_\xi, eff\_e\_\xi\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}$.

@@

\item $A_{\Lambda}$ contains actions of two kinds, like in the original compilation:
\begin{enumerate}[leftmargin=*]
\item Actions for {\em inserting} a {\em precondition}, {\em positive}/{\em negative} effect in $\xi$ following the syntactic constraints of \strips\ models. 
\begin{itemize}[leftmargin=*]
\item A precondition $p$ is inserted in $\xi$ when neither $pre_p$, $eff_p$ exist in $\xi$. In addition, a conditional effect is added now for every {\em schematic mutex} $\tup{p,q}$ s.t. both $p$ and $q$ belong to $\in{\mathcal I}_{\Psi,\xi}$ to ban the insertion of two preconditions that are {\em schematic mutex}.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertPre_{p,\xi}})=&\{\neg pre\_p\_\xi, \neg eff\_p\_\xi,\\
&mode_{insert},\neg mode_{inval}\},\\
\cond(\mathsf{insertPre_{p,\xi}})=&\{\emptyset\}\rhd\{pre\_p\_\xi\},\\
&\{pre\_q\_\xi\}\rhd\{mode_{inval}\}.
\end{align*}
\end{small}

\item Actions which support the addition of a {\em negative}/{\em positive} effect $p\in {\mathcal I}_{\Psi,\xi}$ to the action model $\xi$ are extended with two conditional effects, for every {\em schematic mutex} $\tup{p,q}$ s.t. both $p$ and $q$ belong to $\in{\mathcal I}_{\Psi,\xi}$, to ban the insertion of two positive/negative effects that are {\em schematic mutex}:
\end{itemize}

\begin{small}
\begin{align*}
\pre(\mathsf{insertEff_{p,\xi}})&=\{\neg eff\_p\_\xi, mode_{insert},\neg mode_{inval}\},\\
\cond(\mathsf{insertEff_{p,\xi}})&=\{\emptyset\}\rhd\{eff\_p\_\xi\},\\
&\{pre\_q\_\xi,eff\_q\_\xi,pre\_p\_\xi\}\rhd\{mode_{inval}\},\\
&\{\neg pre\_q\_\xi,eff\_q\_\xi,\neg pre\_p\_\xi\}\rhd\{mode_{inval}\}.
\end{align*}
\end{small}

\item Actions for {\em applying} an action model $\xi$ built by the {\em insert} actions and bounded to objects $\omega\subseteq\Omega^{|pars(\xi)|}$. The action parameters, $pars(\xi)$, are bound to the objects in $\omega$ that appear in the same position. The definition $\mathsf{apply_{\xi,\omega}}$ actions is more compact than the one previously proposed by~\citeauthor{aineto2018learning}~\citeyear{aineto2018learning} since we are not using disjunctions to code the possible preconditions of an action schema. 
\end{enumerate}
\end{itemize}

\begin{small}
\begin{align*}
\pre(\mathsf{apply_{\xi,\omega}})=&\{\neg mode_{inval}\},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{pre\_p\_\xi\wedge eff\_p\_\xi\}\rhd\{\neg p(\omega)\}_{\forall p\in{\mathcal I}_{\Psi,\xi}},\\
&\{\neg pre\_p\_\xi \wedge eff\_p\_\xi\}\rhd\{p(\omega)\}_{\forall p\in{\mathcal I}_{\Psi,\xi}}\},\\
&\{pre\_p\_\xi \wedge \neg p(\omega)\}\rhd\{mode_{inval}\}_{\forall p\in{\mathcal I}_{\Psi,\xi}},\\
&\{\emptyset\}\rhd\{\neg mode_{insert}\},
\end{align*}
\end{small}

For every {\em state-invariant} $\phi\in\Phi$ we can extend $\mathsf{apply_{\xi,\omega}}$ actions with a conditional effect $\{\neg\phi\}\rhd\{mode_{inval}\}$ that checks the consistency of the {\em state-invariant} $\phi$ at every state traversed by a solution to the $P_\Lambda$ problem. Cheking arbitrary $\phi$ formulae can be too expensive for current classical planners. Instead, we prefer to add these conditional effects into $\mathsf{apply_{\xi,\omega}}$ actions for checking the consistency of a {\em schematic mutex} $\tup{p,q}$ s.t. both $p$ and $q$ belong to $\in{\mathcal I}_{\Psi,\xi}$.

\begin{small}
\begin{align*}
&\{pre\_p\_\xi \wedge q(\omega)\}\rhd\{mode_{inval}\}_{\forall p\in{\mathcal I}_{\Psi,\xi}},\\
\end{align*}
\end{small}

\subsection{The bias of the initially {\em empty} action model}
Classical planners tend to preffer shorter solution plans, so our compilation may introduce a bias to $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning tasks preferring solutions that are referred to action models with a shorter number of {\em preconditions}/{\em effects}. In more detail, all $\{pre\_e\_\xi, eff\_e\_\xi\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}$ fluents are false at the initial state of our $P_\Lambda$ compilation so classical planners tend to solve $P_\Lambda$ with plans that require a shorter number of {\em insert} actions.

This bias could be eliminated defining a cost function for the actions in $P_\Lambda$ (e.g. {\em insert} actions have {\em zero cost} while $\mathsf{apply_{\xi,\omega}}$ actions have a {\em positive constant cost}). In practice we use a different approach to disregard the cost of {\em insert} actions because classical planners are not proficiency optimizing {\em plan cost} when there are zero-cost actions. Instead, our approach is to use a SAT-based planner~\cite{rintanen2014madagascar} that can apply all actions for inserting preconditions in a single planning step (these actions do not interact). Further, the actions for inserting action effects are also applied in another single planning step. The plan horizon for programming any action model is then always bound to 2, which significantly reduces the planning horizon. The SAT-based planning approach is also convenient because its ability to deal with classical planning problems populated with dead-ends and because symmetries in the insertion of preconditions/effects into an action model do not affect to the planning performance.

\subsection{Compilation properties}
\begin{mylemma}
Soundness. Any classical plan $\pi_{\Lambda}$ that solves $P_{\Lambda}$ produces a \strips\ model $\mathcal{M'}$ that solves the $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task.
\end{mylemma}

\begin{proof}[Proof]
\begin{small}
According to the $P_{\Lambda}$ compilation, once a given precondition or effect is inserted into the action model $\mathcal{M}$ it cannot be removed back. In addition, once the action model $\mathcal{M}$ is applied it cannot be {\em programmed}. In the compiled planning problem $P_{\Lambda}$, only $\mathsf{apply_{\xi,\omega}}$ actions can update the value of the state fluents $F$. This means that a state consistent with an observation $s_n^o$ can only be achieved executing an applicable sequence of $\mathsf{apply_{\xi,\omega}}$ actions that, starting in the corresponding initial state $s_0^o$, validates that every generated intermediate state $s_i$, s.t. $0\leq i\leq n$, is consistent with the input state observations and {\em state-invariants}. This is exactly the definition of the solution condition for an action model $\mathcal{M}'$ to solve the $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task. 
\end{small}
\end{proof}

\begin{mylemma}
Completeness. Any \strips\ model $\mathcal{M}'$ that solves the $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task can be computed with a classical plan $\pi_{\Lambda}$ that solves $P_{\Lambda}$.
\end{mylemma}

\begin{proof}[Proof]
\begin{small}
By definition ${\mathcal I}_{\Psi,\xi}$ fully captures the set of elements that can appear in a \strips\ action schema $\xi$ using predicates $\Psi$. In addition the $P_{\Lambda}$ compilation does not discard any possible action model $\mathcal{M}'$ definable within ${\mathcal I}_{\Psi,\xi}$ that satisfies the domain mutex in $\Phi$. This means that, for every \strips\ model $\mathcal{M}'$ that solves the $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$, we can build a plan $\pi_{\Lambda}$ that solves $P_{\Lambda}$ by selecting the appropriate $\mathsf{insertPre_{p,\xi}}$ and $\mathsf{insertEff_{p,\xi}}$ actions for {\em programming} the precondition and effects of the corresponding action model $\mathcal{M}'$ and then, selecting the corresponding $\mathsf{apply_{\xi,\omega}}$ actions that transform the initial state observation $s_0^o$ into the final state observation $s_m^o$.
\end{small}
\end{proof}
